# 2: Regression and model validation

*Describe the work you have done this week and summarize your learning.*

## Data wrangling.

After understanding the metadata file and skipping the Finnish language instructions that were unnecessary it was easy to prepare a data file [this data file](https://github.com/crastr/IODS-project/blob/master/data/learning2014.txt). The script can be found [here](https://github.com/crastr/IODS-project/blob/master/data/create_learning2020.R).I used plyr and tiidyverse to do the conversions of the raw data into a workable file.

## Analysis
The dataset is the result of a survey on teaching and learning that was conducted as a research project in 2014.
The dataset was loaded in the next chunk and the dimensions and structure of the document are output.
```{r}
library(tidyverse)
```

```{r}
learning2014_csv <- read_csv("/home/alex/ods_course/IODS-project_R/data/learning2014_AA.csv")
dim(learning2014_csv)

```
```{r}
str(learning2014_csv)

```
The data.frame contains 166 lines and 7 columns. The first column encodes gender, and has the character type. The rest are all numeric, and contain the data which will be used for the regression analysis.
The *"attitude"*, *"deep"*, *"stra"*, and *"surf"* columns contain the combinations of results from the original dataframe. The full metadata explanation can be found [here](https://github.com/crastr/IODS-project/blob/master/data/JYTOPKYS2-meta.txt). The explanation for the study can be found [here](http://www.slideshare.net/kimmovehkalahti/the-relationship-between-learning-approaches-and-students-achievements-in-an-introductory-statistics-course-in-finland).
Briefly:
  - *attitude* variable is *Global attitude toward statistics* composite score of the questions having to do with global attitude toward statistics
  - *deep* variable is *Deep approach* adjusted composite score of the questions connected with deep understanding of statistics
  - *stra* variable is *Strategic approach* adjusted composite score of the questions connected with how strategically the participant approaches the subject
  - *surf* variable is *Surface approach* adjusted composite score of the questions connected with how whether the participant can understand the material deeply and whther the participants has problems studying.

```{r summaries}
summaries <- learning2014_csv %>%
  summarise(across(where(is.numeric), list(mean = ~mean(.), sd = ~sd(.), min = ~min(.), max = ~max(.), median = ~median(.))))

print(t(summaries))
```

```{r}
ggplot(learning2014_csv, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#D6604D", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()

# Boxplot for 'attitude'
ggplot(learning2014_csv, aes(x = deep)) +  # x = 1 creates a single boxplot
  geom_density(fill = "#D6604D", color = "black") +
  labs(title = "Boxplot of Deep approach score", x = "", y = "Attitude") +
  theme_minimal()

# Boxplot for 'attitude'
ggplot(learning2014_csv, aes(x = stra)) +  # x = 1 creates a single boxplot
  geom_density(fill = "#D6604D", color = "black") +
  labs(title = "Boxplot of strategic approach score", x = "", y = "Stra") +
  theme_minimal()

ggplot(learning2014_csv, aes(x = surf)) +  # x = 1 creates a single boxplot
  geom_density(fill = "#D6604D", color = "black") +
  labs(title = "Boxplot of surface approach score", x = "", y = "") +
  theme_minimal()

# For the categorical variable 'gender', a bar chart can be used
ggplot(learning2014_csv, aes(x = points)) +
  geom_bar(fill = "#D6604D", color = "black") +
  labs(title = "Points Distribution", x = "Points", y = "Count") +
  theme_minimal()
```



- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

Here we go again...
